Effectful Semantics for Substructural SSA

### 2023-11-01 abstract-freyd/paper.tex
Began talk abstract

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

@@

\title{Freyd Categories are SSA}

\author{Neel Krishnaswami}
\email{nk480@cl.cam.ac.uk}
\orcid{0000-0003-2838-5865}

@@

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>00000000.0000000.0000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>00000000.00000000.00000000</concept_id>
  <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
\ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}

@@

\TODO{this}

\section{Related Work}

\TODO{\cite{promonad}}

\TODO{\cite{linear-state-usage}}

\TODO{\cite{ssa-is-fun}}

\TODO{\cite{sparky}}

>>>>

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
% \setcopyright{acmcopyright}
% \copyrightyear{2018}
% \acmYear{2018}
% \acmDOI{XXXXXXX.XXXXXXX}

@@

\title{SSA is Freyd Categories}

@@

@@

Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure \cite{llvm}, GCC's GIMPLE IR, Cranelift \cite{cranelift}, and MLIR \cite{mlir} are based. One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses.

One of the interesting observations of the MLIR project is that SSA is very widely applicable, being useful for reasoning about everything from high-level machine learning operations to low-level GPU code, and even hardware design and simulation, as in the CIRCT project \cite{circt}. This hints that SSA is a natural syntax for some fundamental, hopefully compositional, structure, much like how the simply typed lambda calculus is a natural syntax for arbitrary Cartesian closed categories. In this talk, we would like to argue that SSA without control-flow is a natural syntax for \textit{Freyd categories}, which we then provide a graphical syntax for in terms of string diagrams as in \cite{promonad}. We then argue that SSA with \textit{terminating} control-flow (i.e., in which the basic blocks form a DAG) corresponds exactly to \textit{distributive} Freyd categories, while allowing general control-flow corresponds to equipping a distributive Freyd category with an \textit{Elgot structure}; we then give a string-diagrammatic graphical representation of such programs similar to traditional control-flow graphs.

We do all this by giving an inductive, type-theoretic presentation of SSA which we equip with a categorical semantics, proving substitution and the soundness of various basic optimizations and rewrites. Finally, we wish to give some examples of various models and their applications, highlighting how they appear in diverse areas of computer science including classical compiler development, weak memory, quantum circuits, and hardware design.

Time permitting, we also hope to explain how to generalize our IR with support for both substructural types and impure substitutions, which require us to move to the more general setting of \textit{distributive effectful categories}, and give a few applications of this more general system in separation logic and reasoning about nondeterminism.

### 2023-11-01 pldi24/paper.tex
Began talk abstract

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}

@@

Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure, GCC's GIMPLE intermediate representation, and MLIR are based. One of the core advantages of putting programs in SSA form is the ability to reason about programs in an equational manner by 
...

>>>>

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2023}
\acmYear{2023}
\acmDOI{XXXXXXX.XXXXXXX}

@@

Static Single Assignment, or SSA, is a cornerstone of modern compiler design, on which projects such as the LLVM compiler infrastructure, GCC's GIMPLE intermediate representation, Cranelift, and MLIR are based. One of the core advantages of putting programs in SSA form is the ability to reason about programs in a more equational manner: since variables always have exactly one definition, we can justify complex optimizations like global value numbering (which would otherwise need to take into account complex flow-sensitive definitions) and more easily implement various analyses. The semantics of SSA, however, can be quite complex to reason about, especially when considering the somewhat unintuitive scoping rules of variables and \(\phi\)-function arguments.

In this paper, we give a type-theoretic account of SSA-style programs, which we extend with support for substructural types. This allows us to perform proofs and inductive definitions over SSA programs much more easily, which we take advantage of to give SSA a categorical semantics in terms of distributive effectful categories equipped with an Elgot structure. We then give some models, and prove various optimizations sound on them.
